{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-08-24 ms.tex
Minor comments from reviewers

Plant growth and demography are fundamentally size- and 
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
Ontogenetic 
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates 
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

@@

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% 
We developed {\plant} primarily to further investigate if and how species
differing in traits are able to coexist with one another
\citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
provide unique opportunities for coexistence via differentiation in
successional strategy \citep{Moorcroft-2001,Huston-1987}.

@@

Below we describe the general approach of the package, then provide
sections with details at nested levels of ecological organisation.
Detailed technical documents are provided as supplementary information
(see Appendices), updated versions of these technical documents are
available within the {\plant} package itself.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. By being driven by traits, the
model can be extended to potentially very many species. Users will
also be able to add their own physiological models, though this is a
work in progress.

@@

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. To properly model
trait variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs (see Appendix \ref{sec:parameters}).  We refer to
this default model as \texttt{FFW16}, reflecting our initials in this
paper. For example, the trait
\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
rate of leaf turnover, based on an observed scaling relationship
spanning across diverse vegetation and plant functional types
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
some parameters were limited. Providing updated parameter sets is
trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
suitable values by constraining outputs from {\plant} against diverse
data sources \citep{Lebauer-2012, Keenan-2013}. This is an area in
which we are still actively working and one of the major use cases we
envisage for {\plant}.

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
regarded decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

@@

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and sharing a common seed dispersal pool
(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
that the disturbances remove all established vegetation.  With this
approach we can scale up from simulating a single patch to simulate an
entire metapopulation.  The frequency-density \(P(a)\) of patches aged
\(a\) in the landscape is needed. Assuming there are a large number of
patches means the dynamics of \(P\) behave deterministically and can
be approximated via a second \textsc{pde} (see Appendix
\ref{sec:demography} for details). Scaling from patch to
metapopulation is then achieved by weighting the temporal dynamics of
a single patch by \(P(a)\), the relative abundance of patches aged
\(a\) in the metapopulation (Fig. \ref{fig:patch}c).

The main numerical challenge in stepping from a single patch to a
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate the
demography of the metapopulation, this in turn produces an ``output''
number of seeds. The equilibrium occurs when a given input seed rain
produces an equivalent output seed rain. Finding this equilibrium
requires finding the root of a multi-dimensional function, subject to
the constraint that this root is stable (e.g., the trivial equilibrium
of zero seed rain always exists, but is often not stable; Appendix
\ref{sec:equilibrium}).

@@

Finally, with the extension to the metapopulation level, we are
modelling demography across the entire life cycle and are thus in a
position to estimate invasion fitness -- the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities
\citep{Geritz-1998,Dieckmann-2007}.  While ideally calculated as the
long-term per capita growth rate, a more convenient indicator of
fitness in structured metapopulations is given by the
logarithm of the basic reproduction ratio, \(R\)
\citep{Gyllenberg-2001, Metz-2001}.
The basic reproduction ratio is simply the total seed output of 
the new type, averaged across the metapopulation. So a new type can 
invade when \(\log R > 0\).

With this measure of fitness, we can model trait-based adaptive
dynamics and community assembly \citep{Falster-2015}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits
(Fig. \ref{fig:fitness}). The reproductive success of individuals
possessing a particular trait value depends on both the traits and
densities of other species in the current resident community. Fitness
landscapes will be positive in regions that allow invasion, and zero
for species that are at equilibrium. The slope of the fitness
landscape with respect to traits indicates the direction of selection
(dashed line in Fig. \ref{fig:fitness}a) indicates selection for
smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution. Stable coexistence is
possible when species with different traits can mutually invade
\citep{Geritz-1998} Following the fitness gradient may lead to an
evolutionary branching point \citep{Geritz-1998,Dieckmann-2007}; at
this point there is no directional selection (solid line in
Fig. \ref{fig:fitness}a) but invasion is possible both above and below
it in trait space. It is often possible to find a configuration where
fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
except for the zero points where species are present. This community
is immune to invasion from any other strategies and represents a
prediction for the trait mixture that might might be favoured by
natural selection \citep{Geritz-1998,Falster-2015}.

>>>>

Plant growth and demography are fundamentally size- and 
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
Ontogenetic 
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates 
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

@@

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.ƒ
%
We developed {\plant} primarily to further investigate if and how species
differing in traits are able to coexist with one another
\citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep{Moorcroft-2001,Huston-1987}.

@@

Below we describe the general approach of the package, then provide
sections with details at nested levels of ecological organisation.
Detailed technical documents are provided as supplementary information
(see Appendices); updated versions of these technical documents are
available within the {\plant} package itself.

@@

{\plant} is an individual-based model, meaning the dynamics of the population arise from rules specifying how individuals grow and interact. The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. By being driven by traits, the
model can be extended to potentially very many species. Users will
also be able to add their own physiological models, though this is a
work in progress.

@@

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. We refer to
this default model as \texttt{FFW16}, reflecting our initials in this
paper.  To model
trait variation properly, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs (see Appendix \ref{sec:parameters}).  For example, the trait
\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
rate of leaf turnover, based on an observed scaling relationship
spanning across diverse vegetation and plant functional types
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
some parameters were limited. Providing updated parameter sets to the
algorithm is
trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
suitable values by constraining outputs from {\plant} against diverse
data sources \citep{Lebauer-2012, Keenan-2013}. This is an area in
which we are still actively working and one of the major use cases we
envisage for {\plant}.

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
observed decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

@@

In {\plant} and following \citep{Kohyama-1993, Moorcroft-2001,
  Falster-2011} we assume an infinite number of patches, all experiencing
the same disturbance regime and sharing a common seed dispersal pool
(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
that the disturbances remove all established vegetation.  With this
approach we can scale up from simulating a single patch to simulate an
entire metapopulation.  The frequency-density \(P(a)\) of patches aged
\(a\) in the landscape is needed. Assuming there are a large number of
patches means the dynamics of \(P\) behave deterministically and can
be approximated via a second \textsc{pde} (see Appendix
\ref{sec:demography} for details). Scaling from patch to
metapopulation is then achieved by weighting the temporal dynamics of
a single patch by \(P(a)\), the relative abundance of patches aged
\(a\) in the metapopulation (Fig. \ref{fig:patch}c).

The main numerical challenge in stepping from a single patch to a
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate the
demography of the metapopulation, this in turn produces an ``output''
number of seeds. The equilibrium occurs when, for each species,
a given input seed rain
produces an equivalent output seed rain. Finding this equilibrium
requires finding the root of a multi-dimensional function, subject to
the constraint that this root is stable (e.g., the trivial equilibrium
of zero seed rain always exists, but is often not stable; Appendix
\ref{sec:equilibrium}).

@@

Finally, with the extension to the metapopulation level, we are
modelling demography across the entire life cycle and are thus in a
position to estimate invasion fitness -- the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities
\citep{Metz-1992, Geritz-1998}.  While ideally calculated as the
long-term per capita growth rate, a more convenient indicator of
fitness in structured metapopulations is given by the
logarithm of the basic reproduction ratio, \(R\)
\citep{Gyllenberg-2001, Metz-2001}.
The basic reproduction ratio is simply the total seed output of
the new type, averaged across the metapopulation. So a new type can
invade when \(\log R > 0\).

With this measure of fitness, we can model trait-based adaptive
dynamics and community assembly of forest communities, following
 \citep{Falster-2015}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits
(Fig. \ref{fig:fitness}). The reproductive success of individuals
possessing a particular trait value depends on both the traits and
densities of other species in the current resident community. Fitness
landscapes will be positive in regions that allow invasion, and zero
for species that are at equilibrium. The slope of the fitness
landscape with respect to traits indicates the direction of selection
(dashed line in Fig. \ref{fig:fitness}a) indicates selection for
smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution. Stable coexistence is
possible when species with different traits can mutually invade
\citep{Geritz-1998, Chesson-2000}. Following the fitness gradient may lead to an
evolutionary branching point \citep{Geritz-1998}; at
this point there is no directional selection (solid line in
Fig. \ref{fig:fitness}a) but invasion is possible both above and below
it in trait space. It is often possible to find a configuration where
fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
except for the zero points where species are present. This community
is immune to invasion from any other strategies and represents a
prediction for the trait mixture that might might be favoured by
natural selection \citep{Geritz-1998}.
